Co-evolution of lexical meaning \& pragmatic use

### 2017-04-12 paper/paper-v3.tex
sligth draft changes

If this majority view is correct, the question arises how such a division of labor between
semantics and pragmatics could have evolved, why it would be so pervasive across natural
languages and why it is that some expressions systematically draw from it while others semantically conventionalize. Models of language evolution abound. There are simulation-based models studying
populations of communicating agents
\citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
and there are mathematical models of language evolution, mostly coming from game theory
\citep{lewis:1969,Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much
of this work has focused on explaining basic properties such as compositionality and
combinatoriality
\citep[e.g.,][]{Batali1998:Computational-S,nowak+krakauer:1999,nowak+etal:2000,KirbyHurford2002:The-Emergence-o,kirby:2002,SmithKirby2003:Iterated-Learni,Gong2007:Language-Evolut,kirby+etal:2015,verhoef+etal:2014,Franke2015:Proto-Syntax},
but little attention has been paid to the interaction between conventional meaning and
pragmatic use. What is more, many mathematical models explain evolved meaning as a regularity
in the overt behavior of agents, abstracting from complex interactions between semantic
representations and pragmatic use. In contrast, we will here look at language users with a
richer cognitive make-up.

We spell out a model of the co-evolution of conventional meaning and pragmatic reasoning
types. The objects of replication and selection are pairs consisting of a set of lexical
meanings and a manner of pragmatic behavior. The latter are represented using probabilistic
models of pragmatic language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
and selection are described by the \emph{replicator mutator dynamic}, a general and established
model of evolutionary change in large and homogeneous populations
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. This
approach allows us to study the interaction between (i) pressure towards communicative
efficiency and (ii) infidelity in the transmission of linguistic knowledge, caused by factors
such as inductive learning biases and sparse learning data. Considering transmission of
linguistic knowledge is important because neither semantic meanings nor pragmatic usage
patterns are directly observable. Instead, language learners have to infer these unobservables
from the observable behavior in which they result. We formalize this process as a form of
Bayesian inference. Our approach thereby contains as a special case a well-understood model of
iterated Bayesian learning \citep{griffiths+kalish:2007}, but combines it with functional
selection, here formalized as the most versatile dynamic from evolutionary game theory, the
replicator dynamic \citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces
this model.

Section~\ref{sec:si-case-study} applies this model to a case study on scalar implicatures. We
discuss a setting in which the majority view of underspecified lexical meanings and pragmatic
enrichments emerges if selection and transmission infidelity are combined. In particular, we
show that inductive learning biases of Bayesian learners that favor simpler lexical meanings
can prevent the lexicalization of scalar inferences and lead to the emergence of Gricean-like
pragmatic reasoning types. Results of this case study are critically assessed in the light of
the assumptions that feed our model in Section~\ref{sec:discussion}.

@@

The idea that language is an adaptation to serve a communicative function is fundamental to
many synchronic and diachronic analyses at least since \citeposs{zipf:1949} explanation of word
frequency rankings as a result of competing hearer and speaker preferences \citep[e.g.,
in][]{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
  piantadosi:2014,kirby+etal:2015}. If processes of selection, such as conditional imitation or
reinforcement, favor more communicatively efficient types of behavior, languages are driven
towards semantic expressivity \citep[e.g.,][]{nowak+krakauer:1999,Skyrms2010:Signals}. But
pressure towards communicative efficiency is not the only force that shapes
language. Learnability is another as natural languages need to be learnable to survive their
faithful transmission across generations. Furthermore, even small learning biases implicit in acquisition
can build up and have quite striking effects on an evolving language in a process of iterated
learning
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}.

@@

\begin{example}
  Consider a simple coordination game. Agents are of two types: positive or negative. If
  agents of different type play with each other, they obtain a payoff of 0. If negative meets
  negative, each receives a payoff of 1. If positive meets positive, they get a payoff of 2. A
  population state is completely characterized by the proportion $x$ of negatives. The fitness
  of negatives in population state $x$ is $f_n(x) = x$, that of positives is $f_p(x) = 2-2x$.
  The average fitness is $\Phi(x) = x f_n(x) + (1-x) f_p(x) = 3x^2 - 4x + 2$. Without mutation,
  the replicator dynamic will then update $x$ to $RD(x) = \nicefrac{f(x)x}{\Phi(x)} = \nicefrac{x^2}{\Phi(x)}$. The update
  function $RD(x)$ of the replicator step is plotted in Figure~\ref{fig:Updates_RMS} as the
  blue line. Rest points, for which $RD(x)=x$, are at $x=0$, $x=1$ and $x= \nicefrac{2}{3}$.
  The former are attractors as nearby points converge to them. Points near $x=\nicefrac{2}{3}$
  move towards 0 or 1. This is schematically pictured in the topmost phase portrait in
  Figure~\ref{fig:Phase_RD}.

@@

Agents play signaling games, in which the speaker wants to communicate a world state $s$ with a
message $m$ to a hearer who receives $m$ but does not know $s$
\citep[e.g.][]{lewis:1969,Skyrms2010:Signals}. A lexicon associates each message with a
(possibly fuzzy) set of states. A pragmatic type specifies a probabilistic speaker
rule (a probabilistic choice of message for each state) and a probabilistic hearer rule (a
probabilistic choice of state for each message) given a lexicon. There are many ways of making these general
notions more concrete. Here is what we will assume in the remainder of this paper.

@@

      \mygray{\mall}  & 0.35 & 0.65 \\

@@

With low $\lambda$ speakers choose true messages with more slack. Reasoning over this behavior therefore also results in a weaker association of messages with only true states in receivers, but also in a slightly stronger association of $\msome$ with $\ssome$ over $\sall$ for $\Llack$ users because $S_0(\msome | \ssome;\Llack) > S_0(\msome | \sall;\Llack)$. For $\lambda = 20$, there will be less slack in literal speaker behavior:

@@

And accordingly less slack in pragmatic interpretation:

@@

For $\lambda = 20$, pragmatic speaker behavior of level 1 is as follows:

@@

Let $D$ be the set of possible data that learners may be exposed to. This set $D$ contains all
sequences of state-message pairs of length $k$, e.g.,
$\tuple{\tuple{s_1,m_1},\dots , \tuple{s_k,m_k}}$. As $k$ increases, learners have more data to base their inference on and so tend to
recover the true types that generated a given sequence with higher probability. The mutation
matrix $Q$ of the replicator mutator dynamic in (\ref{eq:RMD_discrete}) can then be defined as
follows: $Q_{ji}$ is the probability that a learner acquires type $i$ when learning from an
agent of type $j$. The learner observes a length-$k$ sequence $d$ of state-message pairs, but
the probability $P(d \mid t_j)$ with which sequence $d = \tuple{\tuple{s_1,m_1},\dots , \tuple{s_k,m_k}}$ is observed depends on type $j$'s
behavior:
\begin{align*}
  P(d = \tuple{\tuple{s_1,m_1},\dots , \tuple{s_k,m_k}} \mid t_j) = \prod_{i = 1}^k S_{n_j}(m_i
  \mid s_i; L_{j})\,,
\end{align*}
where, as before, $n_j$ is $j$'s pragmatic reasoning type and $L_j$ is $j$'s lexicon. For a
given observation $d$, the probability of acquiring type $i$ is $F(t_i \mid d)$, so that:
\begin{flalign*}
  Q_{ji} \propto \sum_{d \in D} P(d \mid t_j) F(t_i \mid d)\,.
\end{flalign*}
The acquisition probability $F(t_i \mid
d)$ given datum $d$ is obtained by probability matching $l = 1$ or a tendency towards choosing
the most likely type $l > 1$ from the posterior distribution $P(\cdot \mid d)$ over types given
the data, which is calculated by Bayes' rule:
\begin{flalign*}
  & F(t_i \mid d) \propto P(t_i \mid d)^l \; \text{ and }\\
  & P(t_i \mid d) \propto P(t_i) P(d \mid t_i)\,.
\end{flalign*}

@@

The model has three parameters: $\lambda$ regulates the degree to which pragmatic speakers
choose messages that appear optimal from the point of view of the agent's own utility measure
(which may be unrelated to the expected utility when communicating with a given population);
$k$ is length of observations for each language learner; $l$ regulates where the learners'
inference behavior lies on a spectrum from probability matching to choice of the most likely
parent type.

>>>>

If this majority view is correct, the question arises how such a division of labor between
semantics and pragmatics could have evolved, why it would be so pervasive across natural
languages, and why it is that some expressions systematically draw from it while others semantically conventionalize. Models of language evolution abound. There are simulation-based models studying
populations of communicating agents
\citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
and there are mathematical models of language evolution, mostly coming from game theory
\citep{lewis:1969,Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much
of this work has focused on explaining basic properties such as compositionality and
combinatoriality
\citep[e.g.,][]{Batali1998:Computational-S,nowak+krakauer:1999,nowak+etal:2000,KirbyHurford2002:The-Emergence-o,kirby:2002,SmithKirby2003:Iterated-Learni,Gong2007:Language-Evolut,kirby+etal:2015,verhoef+etal:2014,Franke2015:Proto-Syntax},
but little attention has been paid to the interaction between conventional meaning and
pragmatic use. What is more, many mathematical models explain evolved meaning as a regularity
in the overt behavior of agents, abstracting from complex interactions between semantic
representations and pragmatic use. In contrast, we will here look at language users with a
richer cognitive make-up.

We spell out a model of the co-evolution of conventional meaning and pragmatic reasoning.
The objects of replication and selection are pairs consisting of a set of lexical
meanings and a manner of pragmatic behavior. The latter are represented using probabilistic
models of pragmatic language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
and selection are described by the \emph{replicator mutator dynamic}, a general and established
model of evolutionary change in large and homogeneous populations
\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. This
approach allows us to study the interaction between (i) pressure towards communicative
efficiency and (ii) infidelity in the transmission of linguistic knowledge, caused by factors
such as inductive learning biases and sparse learning data. Considering transmission of
linguistic knowledge is important because neither semantic meanings nor pragmatic usage
patterns are directly observable. Instead, language learners have to infer these unobservables
from the observable behavior in which they result. We formalize this process as a form of
Bayesian inference. Our approach thereby contains as a special case a well-understood model of
iterated Bayesian learning \citep{griffiths+kalish:2007}, but combines it with functional
selection, here formalized as the most versatile dynamic from evolutionary game theory, the
replicator dynamic \citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces
this model.

Section~\ref{sec:si-case-study} applies this model to a case study on scalar implicatures. We
discuss a setting in which the majority view of underspecified lexical meanings and systematic pragmatic
enrichments emerges if selection and transmission infidelity are combined. In particular, we
show that inductive learning biases of Bayesian learners that favor simpler lexical meanings
can prevent the lexicalization of scalar inferences and lead to the emergence of Gricean-like
pragmatic reasoning types. Results of this case study are critically assessed in the light of
the assumptions that feed our model in Section~\ref{sec:discussion}.

@@

The idea that language is an adaptation to serve a communicative function is fundamental to
many synchronic and diachronic analyses at least since \citeposs{zipf:1949} explanation of word
frequency rankings as a result of competing hearer and speaker preferences \citep[e.g.,
in][]{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
  piantadosi:2014,kirby+etal:2015}. If processes of selection, such as conditional imitation or
reinforcement, favor more communicatively efficient types of behavior, languages are driven
towards semantic expressivity \citep[e.g.,][]{nowak+krakauer:1999,Skyrms2010:Signals}. But
pressure towards communicative efficiency is not the only force that shapes
language. Learnability is another, as natural languages need to be learnable to survive their
faithful transmission across generations. Furthermore, even small learning biases implicit in acquisition
can build up and have quite striking effects on an evolving language in a process of iterated
learning
\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}.

@@

\begin{example}
  Consider a simple and abstract coordination game. Agents are of two types: positive or negative. If
  agents of different type play with each other, they obtain a payoff of 0. If negative meets
  negative, each receives a payoff of 1. If positive meets positive, they get a payoff of 2. A
  population state is completely characterized by the proportion $x$ of negatives. The fitness
  of negatives in population state $x$ is $f_n(x) = x$, that of positives is $f_p(x) = 2-2x$.
  The average fitness is $\Phi(x) = x f_n(x) + (1-x) f_p(x) = 3x^2 - 4x + 2$. Without mutation,
  the replicator dynamic will then update $x$ to $RD(x) = \nicefrac{f(x)x}{\Phi(x)} = \nicefrac{x^2}{\Phi(x)}$. The update
  function $RD(x)$ of the replicator step is plotted in Figure~\ref{fig:Updates_RMS} as the
  blue line. Rest points, for which $RD(x)=x$, are at $x=0$, $x=1$ and $x= \nicefrac{2}{3}$.
  The former are attractors as nearby points converge to them. Points near $x=\nicefrac{2}{3}$
  move towards 0 or 1. This is schematically pictured in the topmost phase portrait in
  Figure~\ref{fig:Phase_RD}.

@@

Agents play signaling games, in which the speaker wants to communicate a world state $s$ with a
message $m$ to a hearer who receives $m$ but does not know $s$
\citep[e.g.][]{lewis:1969,Skyrms2010:Signals}. A lexicon associates each message with a
(possibly fuzzy) set of states. A pragmatic behavior specifies a probabilistic speaker
rule (a probabilistic choice of message for each state) and a probabilistic hearer rule (a
probabilistic choice of state for each message) given a lexicon. There are many ways of making these general
notions more concrete. Here is what we will assume in the remainder of this paper.

@@

      \mygray{\mall}  & .35 & .65 \\

@@

With low $\lambda$ speakers choose true messages with more slack. Reasoning over this behavior therefore also results in a weaker association of messages with only true states in receivers, but also in a slightly stronger association of $\msome$ with $\ssome$ over $\sall$ for $\Llack$ users because they reason that $S_0(\msome | \ssome;\Llack) > S_0(\msome | \sall;\Llack)$. For $\lambda = 20$, there will be less slack in literal speaker behavior:

@@

And accordingly less slack in level $1$ pragmatic interpretation:

@@

For $\lambda = 20$, pragmatic speaker behavior of level 1 is instead as follows:

@@

Let $D$ be the set of possible data that learners may be exposed to. This set $D$ contains all
sequences of state-message pairs of length $k$, e.g.,
$\tuple{\tuple{s_1,m_1},\dots , \tuple{s_k,m_k}}$. As $k$ increases, learners have more data to base their inference on and so tend to
recover the true types that generated a given sequence with higher probability. The mutation
matrix $Q$ of the replicator mutator dynamic in (\ref{eq:RMD_discrete}) can then be defined as
follows: $Q_{ji}$ is the probability that a learner acquires type $i$ when learning from an
agent of type $j$. The learner observes a length-$k$ sequence $d$ of state-message pairs, but
the probability $P(d \mid t_j)$ with which sequence $d = \tuple{\tuple{s_1,m_1},\dots , \tuple{s_k,m_k}}$ is observed depends on type $j$'s
linguistic behavior:
\begin{align*}
  P(d = \tuple{\tuple{s_1,m_1},\dots , \tuple{s_k,m_k}} \mid t_j) = \prod_{i = 1}^k S_{n_j}(m_i
  \mid s_i; L_{j})\,,
\end{align*}
where, as before, $n_j$ is $j$'s pragmatic reasoning type and $L_j$ is $j$'s lexicon. For a
given observation $d$, the probability of acquiring type $i$ is $F(t_i \mid d)$, so that:
\begin{flalign*}
  Q_{ji} \propto \sum_{d \in D} P(d \mid t_j) F(t_i \mid d)\,.
\end{flalign*}
The acquisition probability $F(t_i \mid
d)$ given datum $d$ is obtained by probability matching $l = 1$ or a tendency towards choosing
the most likely type $l > 1$ from the posterior distribution $P(\cdot \mid d)$ over types given
the data, which is calculated by Bayes' rule:
\begin{flalign*}
  & F(t_i \mid d) \propto P(t_i \mid d)^l \; \text{ and }\\
  & P(t_i \mid d) \propto P(t_i) P(d \mid t_i)\,.
\end{flalign*}

@@

The model has three parameters: $\lambda$ regulates the degree to which speakers
choose messages that appear optimal from the point of view of the agent's own utility measure
(which may be unrelated to the expected utility when communicating with a given population);
$k$ is length of observations for each language learner; $l$ regulates where the learners'
inference behavior lies on a spectrum from probability matching to acquisition of the most likely
parent type.
